---
title: "Investigating Drought Climatic Aspects In Puerto Rican Tropical Rainforest"
author: "Trung Dam"
date: "2023-08-17"
output: html_document
---

# Introduction

The dynamic of precipitation and temperature has always been an ongoing research topic. This research aims to delve into the interconnection of daily rainfall and occasionally heat shock in two study sites in Puerto Rico: Sabana Field Research Station and Bisley Experimental Watersheds. These two sites are chosen to investigate the discrepancy of the relationship between rainfall and temperature across different topography. Two sites are located in the Luquillo Mountains, rising from the coast to peaks nearly 1100 meters above sea level (González et al., 2013). Average annual temperatures range from 20\u00B0C to 25\u00B0C(Harris et al., 2012, Ramírez, 2017; data set 2).  Sabana Field Research Station is at the elevation of 100 meters above sea level (Carter et al. 2020), whereas Bisley is located 265-456 meters above sea level (Schellekens, 2000). Sabana Field Research Station is recorded to receive about 3500 mm of annual rainfall, and Bisley is recored to range between 3000 to 4000 mm of annual precipitation. It is worth noting that the Luquillo Mountains had suffered a major drought in 2015, Hurrican Maria and Irma in 2017, which played a great role in driving Luquillo's ecosystem through severe disturbances and changes. This research look into the impact of low rainfall and heat shock periods on Sabana Field Research Station's biodiversity. Meteorological and ecological censuses data was analyzed and fit into linear/logistic models to provide insights for this research. 

# Hypothesis

Our hypothesis is there is a positive correlation between low daily rainfall and high heat shock. In higher elevation area of Luquillo Mountains (Bisley Experimental Watersheds), the occurrences of a low rainfall day combined with high temperature tend to be lower compared to lower elevated area (Sabana Field Research Station). As a consequence, Sabana Field Research Station's biodiversity would increase after the major drought and decrease substantially after hurricanes. 

# Data and Methods
## Data Description

1. Weather data

The meteorological data is collected in two study sites: Sabana Field Research Station and Bisley Experimental Watersheds by USDA Forest Service and Tropical Responses to Altered Climate Experiment (TRACE) from 2015 to 2017. The data sets contain two sites’ daily average temperatures and daily max temperatures (in C°), daily total precipitation (in mm), daily relative humidity and daily maximum relative humidity, timestamp, and date stamp. 


The data got loaded and processed in R. We reformatted the timestamp and date for future time series use. NAs, outliers were removed from the data. Temperature values, which are negative or greater than 50 C°, were eliminated. Relative humidity values had to be greater or equal to 63% (SK Brown, 1983). Daily rainfall values that are more than 400 mm were eliminated as the maximum daily rainfall in Sabana Field Research Station was claimed to be 310 mm (SK Brown, 1983). In addition, some of the data were missing because of an electricity shortage during Hurricane Maria and Irma (August - December 2017). We calculated the daily heat index, a measure of the combined effects of temperature and relative humidity. Daily heat intensity was classified according to the Heat Index Chart from the National Weather Service, including four intensity levels: Hot, Extreme Hot, Danger Hot, Extreme Danger Hot. Furthermore, daily rainfall intensity (Low rainfall and High rainfall) was identified by comparing to a threshold, which is the first percentile of the rolling averages of respectively two study sites (0.1 mm). Climate stressors variable was categorized as day with “Low rainfall” and “Danger Hot” or “Extreme Danger Hot”. This variable was further integrated in our statistical test to clarify the impact of drought on Sabana Field Research Station’s biodiversity. 


2. TRACE Seedling and Sapling Census data

The seedling and sapling census data is collected in Sabana Field Research Station by Tropical Responses to Altered Climate Experiment (TRACE) from 2015 to 2018. The data contains one seedling & sapling census each year except for 2015. There were two censuses that year: one was recorded during the major drought and the other was recorded after. All the seedlings and saplings measured are in the hexagon plots, which cover approximately 12 square meters. They scatter in 6 different plots across topography in the study site. 3 of them are control plots whereas the others are warming plots (4 ° C more than ambient temperature). The data sets contain month stamps, type classification, species classification, height (in cm), diameter (in mm), number of leaves, and damage estimation (in %). 


The data got loaded and processed in R. We calculated Shannon D’s biodiversity index of Sabana Field Research Station’s ecosystem over time. This was later used for observing the changes in the plant's decomposition after the major drought in 2015. The percentage of changes of each plant over time were computed to clarify top drought-resistant plants.

## Statistical Testing

The meterological data are fit in a linear model to explore the relationship between rainfall and temperature change in two study sites: Sabana Field Research Station and Bisley Experimental Watersheds. The data sets further got fitted in a logistic model to investigate the probability of having a "climatic stressors" day (a day with low rainfall and extreme heat) in regard to daily temperature, daily precipitation, and different topography.


```{r Loading libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(cowplot)
library(weathermetrics)
library(zoo)
library(SPEI)
library(vegan)
```

```{r Reading in and Cleaning TRACE weather data from 2014 to 2017, echo=FALSE, message=FALSE, warning=FALSE}
weather2017 <- read.csv("TRACE2017.csv")
weather2016 <- read.csv("TRACE2016.csv")
weather2015 <- read.csv("TRACE2015.csv")
weather2014 <- read.csv("TRACE2014.csv")

#Weather 2014
weather2014 <- weather2014[-(1:3),]
colnames(weather2014)[3] <- "TOT_PREP(mm)"
colnames(weather2014)[1] <- "Timestamp"
colnames(weather2014)[5] <- "AirTC_Avg"
colnames(weather2014)[6] <- "Relative Humidity"
colnames(weather2014)[9] <- "AirTC_Max"


weather2014$Timestamp <- mdy_hm(weather2014$Timestamp)
weather2014$month <- month.name[month(weather2014$Timestamp)]
weather2014$year <- year(weather2014$Timestamp)
weather2014$`TOT_PREP(mm)` <- as.numeric(weather2014$`TOT_PREP(mm)`)
weather2014$`TOT_PREP(mm)` <- ifelse(is.na(weather2014$`TOT_PREP(mm)`), 0, weather2014$`TOT_PREP(mm)`)
weather2014$`Relative Humidity` <- as.numeric(weather2014$`Relative Humidity`)
weather2014$AirTC_Avg <- as.numeric(weather2014$AirTC_Avg)
weather2014$AirTC_Max <- as.numeric(weather2014$AirTC_Max)

weather2014 <- weather2014 %>% group_by(month) %>% mutate(total_Prep = sum(`TOT_PREP(mm)`)) %>%
  select(Timestamp, AirTC_Avg, AirTC_Max, `TOT_PREP(mm)`, total_Prep, `Relative Humidity`, month, year) %>% 
  filter(year == 2014)

#Weather 2015
weather2015 <- weather2015[-(1:3),]
colnames(weather2015)[3] <- "TOT_PREP(mm)"
colnames(weather2015)[1] <- "Timestamp"
colnames(weather2015)[5] <- "AirTC_Avg"
colnames(weather2015)[6] <- "Relative Humidity"
colnames(weather2015)[9] <- "AirTC_Max"


weather2015$Timestamp <- mdy_hm(weather2015$Timestamp)
weather2015$month <- month.name[month(weather2015$Timestamp)]
weather2015$year <- year(weather2015$Timestamp)
weather2015$`TOT_PREP(mm)` <- as.numeric(weather2015$`TOT_PREP(mm)`)
weather2015$`TOT_PREP(mm)` <- ifelse(is.na(weather2015$`TOT_PREP(mm)`), 0, weather2015$`TOT_PREP(mm)`)
weather2015$`Relative Humidity` <- as.numeric(weather2015$`Relative Humidity`)
weather2015$AirTC_Avg <- as.numeric(weather2015$AirTC_Avg)
weather2015$AirTC_Max <- as.numeric(weather2015$AirTC_Max)

weather2015 <- weather2015 %>% group_by(month) %>% mutate(total_Prep = sum(`TOT_PREP(mm)`)) %>%
  select(Timestamp, AirTC_Avg, AirTC_Max, `TOT_PREP(mm)`, total_Prep, `Relative Humidity`, month, year) %>% 
  filter(year == 2015)

#Weather 2016
weather2016 <- weather2016[-(1:3),]
colnames(weather2016)[3] <- "TOT_PREP(mm)"
colnames(weather2016)[1] <- "Timestamp"
colnames(weather2016)[5] <- "AirTC_Avg"
colnames(weather2016)[6] <- "Relative Humidity"
colnames(weather2016)[9] <- "AirTC_Max"


weather2016$Timestamp <- mdy_hm(weather2016$Timestamp)
weather2016$month <- month.name[month(weather2016$Timestamp)]
weather2016$year <- year(weather2016$Timestamp)
weather2016$`TOT_PREP(mm)` <- as.numeric(weather2016$`TOT_PREP(mm)`)
weather2016$`TOT_PREP(mm)` <- ifelse(is.na(weather2016$`TOT_PREP(mm)`), 0, weather2016$`TOT_PREP(mm)`)
weather2016$`Relative Humidity` <- as.numeric(weather2016$`Relative Humidity`)
weather2016$AirTC_Avg <- as.numeric(weather2016$AirTC_Avg)
weather2016$AirTC_Max <- as.numeric(weather2016$AirTC_Max)

weather2016 <- weather2016 %>% filter(!is.na(AirTC_Avg)) %>% group_by(month) %>% mutate(total_Prep = sum(`TOT_PREP(mm)`)) %>%
  select(Timestamp, AirTC_Avg, AirTC_Max, `TOT_PREP(mm)`, total_Prep, `Relative Humidity`, month, year)
  

#Weather 2017
weather2017 <- weather2017[-(1:3),]
colnames(weather2017)[3] <- "TOT_PREP(mm)"
colnames(weather2017)[1] <- "Timestamp"
colnames(weather2017)[5] <- "AirTC_Avg"
colnames(weather2017)[6] <- "Relative Humidity"
colnames(weather2017)[9] <- "AirTC_Max"


weather2017$Timestamp <- mdy_hm(weather2017$Timestamp)
weather2017$month <- month.name[month(weather2017$Timestamp)]
weather2017$year <- year(weather2017$Timestamp)
weather2017$`TOT_PREP(mm)` <- as.numeric(weather2017$`TOT_PREP(mm)`)
weather2017$`TOT_PREP(mm)` <- ifelse(is.na(weather2017$`TOT_PREP(mm)`), 0, weather2017$`TOT_PREP(mm)`)
weather2017$`Relative Humidity` <- as.numeric(weather2017$`Relative Humidity`)
weather2017$AirTC_Avg <- as.numeric(weather2017$AirTC_Avg)
weather2017$AirTC_Max <- as.numeric(weather2017$AirTC_Max)

weather2017 <- weather2017 %>% filter(!is.na(AirTC_Avg) & AirTC_Avg > 0  ) %>%  group_by(month) %>% mutate(total_Prep = sum(`TOT_PREP(mm)`)) %>%
  select(Timestamp, AirTC_Avg, AirTC_Max, `TOT_PREP(mm)`, total_Prep, `Relative Humidity`, month, year)

# Merging data from each year from 2014 to 2017 at TRACE & Cleaning new data
weatherTRACE <- bind_rows(weather2014, weather2015, weather2016, weather2017)

weatherTRACE$AirTC_Avg <- ifelse(is.na(weatherTRACE$AirTC_Avg) |weatherTRACE$AirTC_Avg < 0 , 0, weatherTRACE$AirTC_Avg)
weatherTRACE$`Relative Humidity` <- ifelse(is.na(weatherTRACE$`Relative Humidity`) | weatherTRACE$`Relative Humidity` < 0, 0, weatherTRACE$`Relative Humidity`)

colnames(weatherTRACE)[5] = "total_Precip_month"
weatherTRACE <- weatherTRACE %>% group_by(year) %>% mutate(total_Precip_year = sum(`TOT_PREP(mm)`))
```

## Data exploratory analyses
### Sabana Field Research Station
```{r Plotting total precipitation yearly from 2014 to 2017 at TRACE, echo=FALSE, message=FALSE, warning=FALSE }
totPrecipYearly <- unique(weatherTRACE %>% group_by(year) %>% mutate(numDays = n()) %>% select(year, numDays, total_Precip_year)) %>% filter(year != "NA" & year != 2013) %>% mutate(rainfallDaily = total_Precip_year/numDays)
totPrecipYearly$year <- as.factor(totPrecipYearly$year)
  
ggplot(totPrecipYearly, aes(x = factor(year), y = rainfallDaily, fill = year)) + 
  geom_bar(stat = "identity", show.legend = FALSE) + 
  scale_fill_manual(values = c("2014" = "black", "2015" =  "red", "2016" = "black", "2017" = "black", "2018" = "black", "2019" = "black"))+
  ggtitle("Annual daily rainfall average from 2014 to 2017 at Sabana Field Research Station")+
  ylab("Precipitation in mm")+
  xlab("Year")+
  theme_cowplot()
```

#### Discussion

This bar graph highlights the severe impact of droughts on daily average amount of precipitation in Sabana Field Research Station in 2015. It's necessary to note that we chose to plot out the daily rainfall average due to the inconsistent of our meterological data (missing data for days, shortage of electricity)

```{r Time series point graph about temperature change through out the year, echo=FALSE, message=FALSE, warning=FALSE}
weatherTRACE %>% filter(year == 2015) %>% ggplot(aes(x = Timestamp, y = AirTC_Avg)) + geom_point()+
  labs(title = "2015 Daily Average Temperature at Sabana Field Research Station", x = "Time", y = "Temperature in \u00B0C")+
  theme_cowplot()
```

#### Discussion

Delving into the trend of daily temperature in Sabana Field Research Station in 2015, we notice a temperature spike from May to early November. This immense changes clearly demonstrate the impact of the major drought on Luquillo Mountains' daily temperature.

```{r Plotting temperature, total precipitation, and time - Version 2, echo=FALSE, message=FALSE, warning=FALSE}
tempPrecipDaily <- weatherTRACE %>%  select(Timestamp, AirTC_Avg, `TOT_PREP(mm)`) %>% filter(!year%in% c(2013,2019)  & AirTC_Avg != 0 & `TOT_PREP(mm)` > 0 & `TOT_PREP(mm)` < 1000)

tempPrecipDaily %>% ggplot() + 
  geom_bar(mapping = aes(x = Timestamp, y = `TOT_PREP(mm)` * 100 / 900), stat = "identity", colour = gray(0.5), fill = gray(0.5)) + 
  geom_line(mapping = aes(x = Timestamp, y = AirTC_Avg)) + 
  geom_point(mapping = aes(x = Timestamp, y = AirTC_Avg)) + 
  scale_y_continuous(
    name = expression("Temperature \u00B0C"), 
    sec.axis = sec_axis(~ . * 900 / 100 , name = "Precipitation in mm"))+
  theme_cowplot() + 
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Daily average Temperature and Precipitation at Sabana Field Research Station 2014-2017")
```

#### Discussion

This graph portrays the interconnection of temperature and precipitation in Sabana Field Research Station from 2014 to 2015. Indeed, daily temperature visualizes a sine cycle, showing the temperature trend in dry seasons and rainy seasons in the Luquillo Mountains. Consequently, the peak of rainfall usually happens from April to November except for during the major drought 2015. 


```{r Calculating heat index & categorizing hot temp, echo=FALSE, message=FALSE, warning=FALSE}
weatherTRACE <- weatherTRACE %>% 
  mutate(heatIndex = heat.index(t= AirTC_Max, rh = `Relative Humidity`, temperature.metric = "celsius", output.metric = "fahrenheit", round = 0)) %>% 
  mutate(heatStatus = case_when(heatIndex <= 90 ~ "Hot",
                          heatIndex > 90 & heatIndex < 103 ~ "Extreme Hot",
                          heatIndex >= 103 & heatIndex < 124 ~ "Danger Hot",
                          heatIndex >= 124 ~ "Extreme Danger Hot"))

# Add rainfall status 
rainfallThreshold = 0.1
weatherTRACE <- weatherTRACE %>% mutate(rainfallStatus = case_when(`TOT_PREP(mm)` <= rainfallThreshold ~ "Low rainfall",
                                        `TOT_PREP(mm)` > rainfallThreshold ~ "High rainfall")) 

# summary(rainfallRollingData$rolling_avg) # finding the low rainfall threshold for 7-day average
rainfallThreshold7Days = 2.446
rainfallRollingData <- weatherTRACE %>% mutate(rolling_avg = rollmean(`TOT_PREP(mm)`, k = 7, fill = NA, align = "left")) %>% 
  mutate(rainfallStatus = case_when(rolling_avg <= rainfallThreshold7Days ~ "Low rainfall",
                                    rolling_avg > rainfallThreshold7Days ~ "High rainfall")) 

```

```{r Plotting the temperature and rainfall Sabana, echo=FALSE, message=FALSE, warning=FALSE}
weatherTRACE$heatStatus <- factor(weatherTRACE$heatStatus, levels = c("Extreme Danger Hot", "Danger Hot", "Extreme Hot", "Hot"))
weatherTRACE %>%  filter(`TOT_PREP(mm)` < 150) %>% ggplot(aes(y= AirTC_Avg, x = `TOT_PREP(mm)`, color = heatStatus ))+
  geom_point()+
  geom_smooth() +
  labs(title = "Correlation of heat intensity and daily precipitation at Sabana Field Research Station", y= "Temperature in \u00B0C", x = "Precipitation in mm", color = "Heat Intensity")+
  theme_cowplot()
```

#### Discussion

According to CDC heat index chart, there is a total of five stages of heat intensity in descending order: Extreme Danger Hot, Danger Hot, Extreme Hot, Hot, and Normal. Even though the data points are quite variable for Danger Hot and Extreme Hot days, there is an association between temperature and precipitation in the most heat intense days. This observation supports our hypothesis of the positive correlation between low rainfall and high heat shock. Yet, we need to further test the statistical significance of this relationship.


```{r Plotting the temperature and rainfall Sabana 2, echo=FALSE, message=FALSE, warning=FALSE}
# Version 2
weatherTRACE %>% mutate(comboStressors = ifelse(heatStatus == "Extreme Danger Hot" & rainfallStatus == "Low rainfall", "YES", "NO"))%>% 
  ggplot(aes(y= AirTC_Avg, x = `TOT_PREP(mm)`, color = comboStressors ))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~ heatStatus) +
  labs(title = "Correlation of Climate Stressors Occurences and Heat Intensity at Sabana Field Research Station", x = "Precipitation in mm", y = "Temperature in \u00B0C", color = "Climate Stressors")+
  theme_cowplot()
```

#### Discussion

"Climate Stressors" variable (days with low rainfall and intense heat shock) is introduced to further investigate the relationship between daily temperature and rainfall in Sabana Field Research Station. Zooming in the subset of Extreme Danger Hot days, we notice all "Climate Stressors" days appear here, associating with 0 mm of daily precipitation.

### Bisley Experimental Watersheds
```{r Load Bisley data in,  echo=FALSE, message=FALSE, warning=FALSE}
Bisley2015 <- read.csv("Bisley2015.csv")
Bisley2016 <- read.csv("Bisley2016.csv")
Bisley2017 <- read.csv("Bisley2017.csv")
Bisley2018 <- read.csv("Bisley2018.csv")

#Bisley 2015 cleaning
colnames(Bisley2015)[3] <- "TOT_PREP(mm)"
Bisley2015$Timestamp <- dmy(Bisley2015$Timestamp)
Bisley2015$month <- month.name[month(Bisley2015$Timestamp)]

Bisley2015 <- Bisley2015 %>% select(Timestamp, AirTC_Max, AirTC_Min, `TOT_PREP(mm)`, RH_Max, RH_Min, month, year) %>% filter(!is.na(Timestamp))

#Bisley 2016 cleaning
colnames(Bisley2016)[3] <- "TOT_PREP(mm)"
Bisley2016$Timestamp <- mdy(Bisley2016$Timestamp)
Bisley2016$month <- month.name[month(Bisley2016$Timestamp)]

Bisley2016 <- Bisley2016 %>% select(Timestamp, AirTC_Max, AirTC_Min, `TOT_PREP(mm)`, RH_Max, RH_Min, month, year) %>% filter(!is.na(Timestamp))

#Bisley 2017 cleaning
colnames(Bisley2017)[3] <- "TOT_PREP(mm)"
Bisley2017$Timestamp <- mdy(Bisley2017$Timestamp)
Bisley2017$month <- month.name[month(Bisley2017$Timestamp)]

Bisley2017 <- Bisley2017 %>% select(Timestamp, AirTC_Max, AirTC_Min, `TOT_PREP(mm)`, RH_Max, RH_Min, month, year) %>% filter(!is.na(Timestamp))

weatherBisley <- bind_rows(Bisley2015, Bisley2016, Bisley2017) 
weatherBisley <- weatherBisley %>% filter(AirTC_Max > 0 & AirTC_Min > 0 )
```

```{r Integrating heat index and rainfall classification into Bisley,  echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
weatherBisley <- weatherBisley %>% 
  mutate(heatIndex = heat.index(t= AirTC_Max, rh = RH_Max, temperature.metric = "celsius", output.metric = "fahrenheit", round = 0)) %>% 
  mutate(heatStatus = case_when(heatIndex < 80 ~ "Normal",
                          heatIndex <= 90 & heatIndex >= 80 ~ "Hot",
                          heatIndex > 90 & heatIndex < 103 ~ "Extreme Hot",
                          heatIndex >= 103 & heatIndex < 124 ~ "Danger Hot",
                          heatIndex >= 124 ~ "Extreme Danger Hot")) 

summary(weatherBisley$`TOT_PREP(mm)`) # choose the rainfall threshold

# Add rainfall status 
weatherBisley <- weatherBisley %>% mutate(rainfallStatus = case_when(`TOT_PREP(mm)` <= rainfallThreshold ~ "Low rainfall",
                                        `TOT_PREP(mm)` > rainfallThreshold ~ "High rainfall")) 

# Relationship between temp and precip 
weatherBisley$heatStatus <- factor(weatherBisley$heatStatus, levels = c("Extreme Danger Hot", "Danger Hot", "Extreme Hot", "Hot", "Normal"))
```

```{r Relationship between temp and precip Bisley,  echo=FALSE, message=FALSE, warning=FALSE}
weatherBisley %>% filter(heatStatus != "Extreme Danger Hot") %>% ggplot(aes(y = AirTC_Max, x = `TOT_PREP(mm)`, color = heatStatus))+
  geom_point()+
  geom_smooth()+
  labs(title = "Correlation of heat intensity and daily precipitation at Bisley Experimental Watershed", y= "Temperature in \u00B0C", x = "Precipitation in mm", color = "Heat Intensity")+
  theme_cowplot()
```

#### Discussion

This graph captures the correlation of heat intensity (from daily maximum temperature) and daily precipitation at Bisley Experimental Watersheds. With the higher elevation above the sea level, the daily heat of Bisley tends to be more mild compared to Sabana's. However, similar to Sabana Field Research Station, despite the variability of data points for Extreme Hot and Hot days in Bisley, there is an association between temperature and precipitation in the most heat intense day, supporting our first hypothesis. 

```{r Plotting the temperature and rainfall Bisley 2, echo=FALSE, message=FALSE, warning=FALSE }
# Version 2
weatherBisley %>% mutate(comboStressors = ifelse(heatStatus == "Extreme Danger Hot" | heatStatus == "Danger Hot" & rainfallStatus == "Low rainfall", "YES", "NO")) %>% filter(heatStatus!= "Extreme Danger Hot") %>% 
  ggplot(aes(y= AirTC_Max, x = `TOT_PREP(mm)`, color = comboStressors ))+
  geom_point()+
  geom_smooth()+
  facet_wrap(~ heatStatus)+
  labs(title = "Correlation of Climate Stressors Occurences and Heat Intensity at Bisley Experimental Wateshed", x = "Precipitation in mm", y = "Temperature in \u00B0C", color = "Climate Stressors")+
  theme_cowplot()
```

#### Discussion

"Climate Stressors" variable (days with low rainfall and intense heat shock) is introduced to further investigate the relationship between daily temperature and rainfall in Bisley Experimental Watersheds. Looking at the subset of Extreme Danger Hot days, we notice all "Climate Stressors" days appear here, receiving 0 mm of daily precipitation. This further supports our hypothesis of relationship between low rainfall and heat shock.

### Sabana Field Research Station vs. Bisley Experimental Watersheds

```{r Comparison between TRACE and Bisley, echo=FALSE, message=FALSE, warning=FALSE }
# Compare rainfall from two sites

rainfall2sites <- rbind(weatherTRACE %>% group_by(year) %>% summarize(avgPrecip_Yearly = mean(`TOT_PREP(mm)`), Station = "Sabana"), weatherBisley %>% group_by(year) %>% summarize(avgPrecip_Yearly = mean(`TOT_PREP(mm)`), Station = "Bisley")) %>% filter(year != 2019)

rainfall2sites %>% ggplot(aes(x=year, y = avgPrecip_Yearly, fill = Station))+
  geom_bar(stat = "identity", position = "dodge")+
  geom_text(aes(label = round(avgPrecip_Yearly, 1)), 
            position = position_dodge(0.9),
            color="white",vjust = 1,hjust = 0.5)+
  labs(title = "Daily Average of Precipitation in mm in Bisley and Sabana from 2015 to 2017",
       x = "Year",
      y = "Daily Average of Precipitation in mm")+
  theme_cowplot()
```

#### Discussion

From 2014 to 2017, there is a decrease in total precipitation yearly in both Bisley and TRACE Research Stations. In these four years, Bisley had more rainfall amount compared to TRACE, suggesting a difference stem from their ecological attributes. Indeed, higher elevation areas are more likely to receive more direct rain because of less amount of canopy area. It is worth noting that we do not have rainfall data of Bisley in hand, which answers to the lack of a column in 2014 from Bisley.

```{r Compare two sites temperature, echo=FALSE, message=FALSE, warning=FALSE }
# hist(weatherTRACE$AirTC_Max)
# hist(weatherBisley$AirTC_Max) 

inner_join(weatherTRACE %>% select(Timestamp, AirTC_Max), weatherBisley %>% select(Timestamp, AirTC_Max), by = "Timestamp") %>% 
  ggplot(aes(x=Timestamp))+
  geom_line(aes(y = AirTC_Max.x, color = "Sabana")) +
  geom_line(aes(y = AirTC_Max.y, color = "Bisley")) +
  labs(title = "Maximum Temperature in \u00B0C in Bisley and Sabana from 2015 to 2017",
       y = "Temperature in \u00B0C",
       x = "Year",
       color = "Station")+
  theme_cowplot()
```

#### Discussion

Through out 2015 to 2017, Bisley and TRACE had the similar pattern of changes in maximum temperature. However, TRACE had a higher temperature range through out these 3 years, ranging from 22.41 to 34.56 \u00B0C compared to Bisley's temperature, which ranges from 20 to 32.28 \u00B0C. Again, this suggest the difference in topography contributes to different climatic attributes especially hotter weather in lower elevation in this case.

```{r Linear regression model for temperature and rainfall in two sites}
model <- lm(weatherTRACE$`TOT_PREP(mm)`~ AirTC_Max + weatherTRACE$heatStatus + weatherTRACE$rainfallStatus, data = weatherTRACE)
summary(model)

model1 <- lm(weatherBisley$`TOT_PREP(mm)`~ AirTC_Max + weatherBisley$heatStatus + weatherBisley$rainfallStatus , data = weatherBisley)
summary(model1)

```

#### Models Interpretation

These two are two linear regression models between the daily precipitation and daily maximum temperature in two study sites, Bisley and TRACE. For the model in TRACE, with an increase in one Celsius degree, there would be a decrease of 2.3mm in precipitation that day. Similarly for Bisley, with an increase in temperature, there would be a decrease of 0.8mm in precipitation that day. Statistically, two models have low p-values (respectively < 2.2e-16 and 6.059e-12) though insignificant R-squared (respectively 0.145 and 0.15).

```{r Logistic regression models for climate stressors and temp & rainfall in two sites,  echo=FALSE, message=FALSE, warning=FALSE }
# 
# 
# model2 <- glm(weatherTRACE$`Climate Stressors` ~ weatherTRACE$`TOT_PREP(mm)` + AirTC_Max + weatherTRACE$`TOT_PREP(mm)`*AirTC_Max, data = weatherTRACE, family = "binomial")
# summary(model2)
# exp(coef(model2))
# 
# model3 <- glm(weatherBisley$`Climate Stressors` ~ weatherBisley$`TOT_PREP(mm)` + AirTC_Max + weatherBisley$`TOT_PREP(mm)`*AirTC_Max, data = weatherBisley, family = "binomial")
# summary(model3)
# exp(coef(model3))
```

```{r Linear regression models for temperature and rainfall in both sites}
weatherTRACE <- weatherTRACE %>% mutate("Climate Stressors" = ifelse(heatStatus == "Extreme Danger Hot"  & rainfallStatus == "Low rainfall", 1, 0), Station = "TRACE")

weatherBisley <- weatherBisley %>% mutate("Climate Stressors" = ifelse(heatStatus == "Extreme Danger Hot" | heatStatus == "Danger Hot" & rainfallStatus == "Low rainfall", 1, 0), Station = "Bisley")
colnames(weatherBisley)[5] <- "Relative Humidity"

weatherAll <- rbind(weatherTRACE %>% select(- c(AirTC_Avg, total_Precip_month, total_Precip_year)), weatherBisley %>% select(-c(AirTC_Min, RH_Min)))

model4 <- lm(weatherAll$`TOT_PREP(mm)` ~ AirTC_Max + weatherAll$heatIndex + weatherAll$rainfallStatus + weatherAll$Station, data = weatherAll)
summary(model4)

```

#### Model Intepretation

Meteorological data from two sites are merged together. We then fit the data into a linear regression model. With an increase in one Celsius degree, there would be a decrease of 2.1mm in precipitation that day. Statistically, this model has statistical significant p-values (< 2.2e-16) though insignificant R-squared (~0.12)

```{r Comparing frequency of climate stressors in two sites , echo=FALSE, message=FALSE, warning=FALSE}
stressorFreq <- rbind(weatherTRACE %>% group_by(year) %>% count(`Climate Stressors`) %>% mutate(Station = "Sabana"),weatherBisley %>% group_by(year) %>% count(`Climate Stressors`) %>% mutate(Station = "Bisley")) %>% 
  filter(year != 2014 & year != 2019 ) %>% 
  pivot_wider(names_from = `Climate Stressors`, values_from = n)

stressorFreq$`0` <- as.numeric(stressorFreq$`0`)
stressorFreq$`1` <- as.numeric(stressorFreq$`1`)

stressorFreq %>% group_by(year, Station) %>% 
  mutate(yesPer = 100 * `1`/(`1`+`0`)) %>% 
  ggplot(aes(x= year, y = yesPer, fill = Station))+ 
  geom_bar(stat = "identity", position = "dodge")+
  geom_text(aes(label = round(yesPer, 1)), 
            position = position_dodge(0.9),
            color="white",vjust = 1,hjust = 0.5)+
  labs(title = "Frequencies ratio of climate stressors in Bisley and Sabana from 2015 to 2017",
       x = "Year",
      y = "Percentage of having climate stressors yearly (%)")+
  theme_cowplot()

```

#### Discussion

From 2015 to 2017, the percentage of having a climate-stressors day of TRACE is always higher than Bisley's, suggesting that TRACE will be likely to experience low rainfall and high heat shock compared to Bisley.

### Seedling and Sapling Census
```{r Data processing and cleaning, echo=FALSE, message=FALSE, warning=FALSE}
seedlingSapling2015 <- read.csv("SeedlingSapling2015.csv")
seedlingSapling2016 <- read.csv("SeedlingSapling2016.csv")
seedlingSapling2017 <- read.csv("SeedlingSapling2017.csv")

seedlingSapling2015[seedlingSapling2015=='']<- NA
seedlingSapling2015 <- seedlingSapling2015 %>% filter(!is.na(SPECIES))

seedlingSapling2016[seedlingSapling2016=='']<- NA
seedlingSapling2016 <- seedlingSapling2016 %>% filter(!is.na(SPECIES))

seedlingSapling2017[seedlingSapling2017=='']<- NA
seedlingSapling2017 <- seedlingSapling2017 %>% filter(!is.na(SPECIES))


data <- unique(seedlingSapling2015 %>% filter(MONTH == "JUNE") %>%  filter(!grepl("sp|Unknown", SPECIES)) %>% count(SPECIES) %>% mutate(ShannonIndex = diversity(n, "shannon"), year = "Drought 2015") %>% select(year, ShannonIndex))

data1 <- unique(seedlingSapling2015 %>% filter(MONTH == "NOVEMBER") %>%  filter(!grepl("sp|Unknown", SPECIES)) %>% count(SPECIES) %>% mutate(ShannonIndex = diversity(n, "shannon"), year = "Post-drought 2015") %>% select(year, ShannonIndex))

data2 <- unique(seedlingSapling2016 %>% filter(!grepl("sp|UN|Un", SPECIES)) %>% count(SPECIES) %>% mutate(ShannonIndex = diversity(n, "shannon"), year = "2016") %>% select(year, ShannonIndex))

data3 <- unique(seedlingSapling2017 %>%  filter(!grepl("sp|UN|Un", SPECIES)) %>% count(SPECIES) %>% mutate(ShannonIndex = diversity(n, "shannon"), year = "2017" ) %>% select(year, ShannonIndex))
```

```{r Shannon Comparison after drought, echo=FALSE, message=FALSE, warning=FALSE }
ShannonComparison <- rbind(data, data1, data2, data3)
drought2015 = 2.509188
level_order_Shannon <- c("Drought 2015", "Post-drought 2015", "2016", "2017")
ShannonComparison$year <- factor(ShannonComparison$year, levels = c('Drought 2015', "Post-drought 2015", "2016", "2017"))
  ShannonComparison %>% mutate(index_dif = round(ShannonIndex - drought2015, 3)) %>% 
  ggplot(aes(x= factor(year, level= level_order_Shannon), y = index_dif, fill = year)) +
  geom_bar(stat = "identity")+
    labs(title = "Biodiversity Index Difference after major drought 2015 in Sabana Field Research Station", x= "Timeline", y = "Index difference", fill = "Year")+
    theme_cowplot()+
   theme(axis.text.x = element_text(angle = 10, vjust = 0.5, hjust=1))
```

#### Discussion

This graph captures the changes in Shannon biodiversity index of Sabana Field Research Station after the major drought 2015. The result was surprising as our initial hypothesis was after drought 2015, the plant population in Saban would decrease immensely, before bouncing back in 2016 and 2017. However, from Sabana's plant census, the biodiversity of the forest gradually decreases after the major drought 2015. This could be explained by the massive increase of drought-resistance plants in 2015. In 2016 and 2017, when the physiology of the forest stabilized, those plants couldn't compete/ adapt to the restored forest's conditions. That might lead to the extinct of species in Sabana, resulting in lower biodiversity index.

```{r Testing relationship of climate stressors and TRACE biodiversity, echo=FALSE, message=FALSE, warning=FALSE}
climateStressorsvShannon <- weatherTRACE %>% mutate(yearR = case_when(year != 2015 ~ as.character(year),
                                          year == 2015 & month %in% c("January", "February", "March", "April") ~ "Pre-drought 2015",
                                          year == 2015 & month %in% c("May", "June", "July", "August", "September", "October") ~ "Drought 2015",
                                          year == 2015 & month %in% c("November", "December") ~ "Post-drought 2015")) %>% 
                filter(`Climate Stressors` == 1) %>% group_by(yearR)  %>%  count()  %>%  right_join(ShannonComparison, by = c("yearR" = "year")) 
climateStressorsvShannon$n <- climateStressorsvShannon$n %>% replace_na(7) 

climateStressorsvShannon$yearR <- factor(climateStressorsvShannon$yearR, levels = c('Drought 2015', "Post-drought 2015", "2016", "2017"))

climateStressorsvShannon %>% ggplot(aes(x = factor(yearR, levels = level_order_Shannon), y = n, fill = yearR)) +
  geom_bar(stat = "identity")+
    labs(title = "Climate Stressors Occurences overtime at Sabana Reasearch Station", x= "Timeline", y = "Number of days", fill = "Year")+
  theme_cowplot()+
   theme(axis.text.x = element_text(angle = 10, vjust = 0.5, hjust=1))

```

#### Discussion

We looked further into the occurrences of "Climate Stressors" days in Sabana Field Research Station from 2015 to 2017. Obviously, during drought(May-November 2015), there were more "Climate Stessors" day. We then questioned whether if there is a relationship between the number of "Climate Stressors" days and the change of biodiversity index. There were no statistical significance supports the correlation.

```{r ANOVA test}
climatestressorsbyStation <- weatherAll %>% mutate(yearR = case_when(year != 2015 ~ as.character(year),
                                          year == 2015 & month %in% c("January", "February", "March", "April") ~ "Pre-drought 2015",
                                          year == 2015 & month %in% c("May", "June", "July", "August", "September", "October") ~ "Drought 2015",
                                          year == 2015 & month %in% c("November", "December") ~ "Post-drought 2015")) %>% 
                filter(`Climate Stressors` == 1) %>% group_by(yearR, Station)  %>%  count() %>% filter(yearR != 2014)

res.aov <- aov(n ~ Station, data = climatestressorsbyStation)
summary(res.aov)
```

```{r Looking into 2015 - year of drought seedling census, echo=FALSE, message=FALSE, warning=FALSE}
seedlingJune <- seedlingSapling2015 %>% filter(MONTH == "JUNE" & TYPE == "Seedling" & STATUS == "A") %>% select(QUAD, PLOT, TAG, SPECIES, HEIGHT) %>% mutate(Time = "Drought 2015")
seedlingNov <- seedlingSapling2015 %>% filter(MONTH == "NOVEMBER" & TYPE == "Seedling"& STATUS == "A") %>% select(QUAD, PLOT, TAG, SPECIES, HEIGHT) %>% mutate(Time = "Post-drought 2015")

seedlingJune %>% rbind(seedlingNov) %>% ggplot()+
    geom_boxplot(aes(x = SPECIES, y = HEIGHT, fill = Time))+
    labs(title = "Species growth during and after the major drought 2015 in Sabana Field Research Station", x= "Species", y = "Height in cm")+
  theme_cowplot()+
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### Discussion

This visualize investigate the changes in heights of all species that got recorded in Sabana Field Research Station during and after drought. We can notice some of the good drought-resistance species in this forest, including PREMON, PIPHIS, or ROYBOR. We were then eager to capture the species' percent of growth after the major drought.

```{r Percent change of species , echo=FALSE, message=FALSE, warning=FALSE}
seedlingJune %>% left_join(seedlingNov, by = c('QUAD', "PLOT", "TAG", "SPECIES")) %>% mutate(percentChange = (HEIGHT.y*100/HEIGHT.x)-100) %>% 
  ggplot(aes(x = SPECIES, y = percentChange)) +
  geom_boxplot()+
  labs(title = "Species' percentage of growth after the major drought 2015 in Sabana Field Research Station", x= "Species", y = "Height in cm")+
  theme_cowplot()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

seedlingStatTest <- seedlingJune %>% left_join(seedlingNov, by = c('QUAD', "PLOT", "TAG", "SPECIES")) %>% mutate(percentChange = (HEIGHT.y*100/HEIGHT.x)-100) %>% select(TAG, SPECIES, percentChange)
```

#### Discussion

Similar to previous visualization, PREMON, PIPHIS, and ROYBOR are standing out to be one of the best drought-resistance species in this forest. This might suggest residents near Luquillo Mountains to grow them to avoid land fragment during severe drought in the future.

```{r Gibberish, echo=FALSE, message=FALSE, warning=FALSE }
unique( weatherBisley %>% group_by(year) %>% summarize(x = sum(`TOT_PREP(mm)`)))
sum(Bisley2017$`TOT_PREP(mm)`)
unique(weatherTRACE %>% select(year, total_Precip_year))
```